---
title: "Exploring Dengue in Sri Lanka"
author: "Shenan Nathaniel"
format: dashboard
editor: source
theme: flatly
---

```{r}
#| label: load-packages
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(plotly)
library(dygraphs)
library(xts)
library(tsibble)
library(feasts)
library(fabletools)
library(slider)
library(DT)
library(leaflet)
library(denguedatahub)
```

```{r}
#| label: load-data
#| message: false
#| warning: false

data(srilanka_weekly_data)
```

```{r}
#| label: clean-data
#| message: false
#| warning: false

sl <- srilanka_weekly_data |>
  mutate(start.date = parse_date_time(start.date, orders = c("mdy", "ymd"))) |>
  mutate(start.date = as.Date(start.date)) |>
  mutate(epi_week = paste0(year(start.date), " - W", sprintf("%02d", isoweek(start.date)))) |>
  arrange(district, start.date)

sl_total <- sl |>
  group_by(start.date) |>
  summarise(
    cases    = sum(cases, na.rm = TRUE),
    epi_week = first(epi_week),
    .groups  = "drop"
  )
sl_xts <- xts(sl_total$cases, order.by = sl_total$start.date)

annual_district <- sl |>
  mutate(year = year(start.date)) |>
  group_by(district, year) |>
  summarise(annual_cases = sum(cases, na.rm = TRUE), .groups = "drop")

top10 <- annual_district |>
  group_by(district) |>
  summarise(total = sum(annual_cases), .groups = "drop") |>
  slice_max(total, n = 10) |>
  pull(district)

district_coords <- tibble(
  district = c("Colombo", "Gampaha", "Kalutara", "Kandy", "Matale",
               "NuwaraEliya", "Galle", "Matara", "Hambanthota",
               "Jaffna", "Kilinochchi", "Mannar", "Vavuniya", "Mullaitivu",
               "Batticaloa", "Ampara", "Kalmune", "Trincomalee",
               "Kurunegala", "Puttalam", "Anuradhapura", "Polonnaruwa",
               "Badulla", "Monaragala", "Ratnapura", "Kegalle"),
  lat = c(6.927, 7.084, 6.585, 7.291, 7.470, 6.970, 6.054, 5.948, 6.124,
          9.661, 9.395, 8.978, 8.751, 9.267, 7.717, 7.284, 7.414, 8.587,
          7.486, 8.036, 8.335, 7.939, 6.989, 6.873, 6.680, 7.253),
  lon = c(79.861, 80.007, 79.960, 80.636, 80.624, 80.783, 80.221, 80.536, 81.119,
          80.025, 80.399, 79.904, 80.497, 80.812, 81.700, 81.672, 81.836, 81.234,
          80.362, 79.840, 80.406, 81.000, 81.056, 81.349, 80.399, 80.349)
)

yearly_map_data <- left_join(district_coords, annual_district, by = "district")

districts <- sort(unique(sl$district))
```

# Overview

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Total Reported Cases"
list(
  icon  = "virus",
  color = "danger",
  value = format(sum(sl_total$cases, na.rm = TRUE), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Peak Weekly Cases"
peak <- sl_total |> slice_max(cases, n = 1, with_ties = FALSE)
list(
  icon  = "arrow-up-circle",
  color = "warning",
  value = format(peak$cases, big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Peak Epi Week"
list(
  icon  = "calendar-week",
  color = "info",
  value = peak$epi_week
)
```

```{r}
#| content: valuebox
#| title: "Districts Covered"
list(
  icon  = "map",
  color = "success",
  value = n_distinct(sl$district)
)
```

## Row {height="80%"}

### Column {width="100%"}

#### About This Dashboard

This dashboard visualises weekly dengue surveillance data across all 26 administrative districts of Sri Lanka, sourced from the denguedatahub R package, which aggregates reports published by the Epidemiology Unit of the Ministry of Health Sri Lanka. All temporal metrics use Epidemiological Weeks (ISO week standard) rather than calendar dates, consistent with public health reporting conventions.

#### Key Findings

**The 2017 Epidemic**
Sri Lanka's worst dengue outbreak on record occurred in 2017, with 186,101 confirmed cases and 440 deaths reported by the Central Epidemiology Unit — more than three times the annual average of the preceding five years. Cases peaked during the southwest monsoon in June–July 2017, with over 40,000 cases in July alone. The outbreak followed severe flooding in May 2017 that displaced over 600,000 people and created widespread mosquito breeding conditions. *(Source: Ghouse et al., Emerging Infectious Diseases, 2020)*

**Western Province Dominance**
Approximately 43–44% of all dengue cases nationally are concentrated in the Western Province (Colombo, Gampaha, Kalutara). Colombo district alone recorded 23,574 cases in the first seven months of 2017. This reflects both high population density and urban vector breeding conditions. *(Source: ReliefWeb / WHO Sri Lanka, 2017)*

**Seasonal Transmission Pattern**
Dengue transmission in Sri Lanka follows a bimodal seasonal pattern tied to the two monsoon seasons — the southwest monsoon (May–August) and the northeast monsoon (November–January) — with the southwest monsoon consistently driving the larger peak.

*Navigate the tabs above to explore national trends, geographic distribution, district-level time series, and the full data repository.*

# National Trends

## Row

### Column {width="60%"}

```{r}
#| title: "Weekly Transmission Volume — National Total"
dygraph(sl_xts) |>
  dyRangeSelector() |>
  dyOptions(
    strokeWidth = 1.5,
    fillGraph   = TRUE,
    fillAlpha   = 0.12,
    colors      = "#c0392b"
  ) |>
  dyHighlight(highlightSeriesOpts = list(strokeWidth = 2.5))
```

### Column {width="40%"}

```{r}
#| title: "Annual Cases — Top 10 Highest-Burden Districts"
#| message: false
#| warning: false

p_top10 <- annual_district |>
  filter(district %in% top10) |>
  mutate(district = factor(district, levels = top10)) |>
  ggplot(aes(
    x      = year,
    y      = annual_cases,
    colour = district,
    group  = district,
    text   = paste0(
      "<b>", district, "</b><br>",
      "Year: ", year, "<br>",
      "Cases: ", format(annual_cases, big.mark = ",")
    )
  )) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  scale_colour_brewer(palette = "Paired") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Annual Cases", colour = NULL) +
  theme_bw(base_size = 11) +
  theme(
    legend.position  = "bottom",
    legend.text      = element_text(size = 8),
    legend.key.width = unit(1.2, "cm")
  ) +
  guides(colour = guide_legend(nrow = 2))

ggplotly(p_top10, tooltip = "text") |>
  layout(
    legend = list(
      orientation = "h",
      x = 0, y = -0.2,
      font = list(size = 10)
    ),
    hovermode = "closest",
    margin = list(b = 80)
  )
```

# District Map

## Row {.tabset}

### Map by Year

```{r}
#| title: "Geographic Spread by Selected Year"
#| message: false
#| warning: false

max_cases  <- max(yearly_map_data$annual_cases, na.rm = TRUE)
pal        <- colorNumeric(palette = "Reds", domain = c(0, max_cases))
years_list <- sort(unique(yearly_map_data$year))

m <- leaflet() |>
  addTiles() |>
  setView(lng = 80.7, lat = 7.8, zoom = 7)

for (yr in years_list) {
  d <- yearly_map_data |> filter(year == yr)
  m <- m |>
    addCircleMarkers(
      data        = d,
      lng         = ~lon,
      lat         = ~lat,
      radius      = ~pmax(sqrt(annual_cases) / 4, 3),
      color       = ~pal(annual_cases),
      fillColor   = ~pal(annual_cases),
      fillOpacity = 0.85,
      opacity     = 1,
      popup       = ~paste0(
        "<b>District:</b> ", district, "<br>",
        "<b>Year:</b> ", year, "<br>",
        "<b>Annual Cases:</b> ", format(annual_cases, big.mark = ",")
      ),
      label = ~district,
      group = as.character(yr)
    )
}

m |>
  addLayersControl(
    baseGroups = as.character(years_list),
    options    = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    pal      = pal,
    values   = c(0, max_cases),
    title    = "Annual Cases",
    position = "bottomleft"
  )
```

### Annual Case Matrix

```{r}
#| title: "District × Year Case Intensity"
p_matrix <- ggplot(
  annual_district,
  aes(
    x    = year,
    y    = reorder(district, annual_cases, sum),
    fill = annual_cases,
    text = paste0(
      "<b>", district, "</b><br>",
      "Year: ", year, "<br>",
      "Cases: ", format(annual_cases, big.mark = ",")
    )
  )
) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_distiller(
    palette   = "YlOrRd",
    direction = 1,
    name      = "Cases",
    labels    = scales::comma
  ) +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 8)) +
  labs(x = "Year", y = NULL) +
  theme_minimal(base_size = 11) +
  theme(
    axis.text.y     = element_text(size = 9),
    axis.text.x     = element_text(size = 9),
    legend.position = "right"
  )

ggplotly(p_matrix, tooltip = "text") |>
  layout(margin = list(l = 110, r = 20, t = 20, b = 50))
```

# District Overview

## Row

```{r}
#| title: "Weekly Dengue Cases by District — All Sri Lanka"
#| message: false
#| warning: false

sl |>
  ggplot(aes(x = start.date, y = cases)) +
  geom_line(colour = "#c0392b", linewidth = 0.3, alpha = 0.8) +
  facet_wrap(
    ~district,
    ncol   = 5,
    scales = "free_y"
  ) +
  scale_x_date(date_breaks = "4 years", date_labels = "%Y") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Weekly Cases") +
  theme_bw(base_size = 8) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1, size = 6),
    axis.text.y      = element_text(size = 6),
    strip.text       = element_text(face = "bold", size = 7),
    panel.spacing    = unit(0.4, "lines")
  )
```

# District Distributions

## Row {height="100%"}

```{r}
#| title: "Weekly Case Distribution — Top 10 Districts (Violin + Boxplot)"
#| message: false
#| warning: false

p_violin <- sl |>
  filter(district %in% top10) |>
  mutate(district = reorder(district, cases, median)) |>
  ggplot(aes(
    x    = district,
    y    = cases,
    fill = district,
    text = paste0("<b>", district, "</b><br>Cases: ", cases)
  )) +
  geom_violin(alpha = 0.6, linewidth = 0.3) +
  geom_boxplot(
    width        = 0.12,
    outlier.size = 0.8,
    fill         = "white",
    alpha        = 0.8
  ) +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(labels = scales::comma) +
  labs(x = NULL, y = "Weekly Cases") +
  theme_bw(base_size = 11) +
  theme(
    legend.position = "none",
    axis.text.x     = element_text(angle = 30, hjust = 1, size = 10)
  )

ggplotly(p_violin, tooltip = "text") |>
  layout(showlegend = FALSE)
```

# Interactive Trends

## Row

```{r}
#| title: "Weekly Case Trajectory by District"
#| message: false
#| warning: false

traces <- lapply(seq_along(districts), function(i) {
  d <- sl |> filter(district == districts[i])
  list(
    x         = d$start.date,
    y         = d$cases,
    type      = "scatter",
    mode      = "lines",
    name      = districts[i],
    visible   = if (i == 1) TRUE else FALSE,
    text      = paste0(
      "<b>", d$district, "</b><br>",
      "Epi Week: ", d$epi_week, "<br>",
      "Cases: ", d$cases
    ),
    hoverinfo = "text",
    line      = list(color = "#2980b9", width = 1.2)
  )
})

buttons <- lapply(seq_along(districts), function(i) {
  vis    <- rep(FALSE, length(districts))
  vis[i] <- TRUE
  list(
    method = "update",
    args   = list(list(visible = as.list(vis))),
    label  = districts[i]
  )
})

p <- plot_ly()
for (tr in traces) {
  p <- add_trace(
    p,
    x         = tr$x,
    y         = tr$y,
    type      = tr$type,
    mode      = tr$mode,
    name      = tr$name,
    visible   = tr$visible,
    text      = tr$text,
    hoverinfo = tr$hoverinfo,
    line      = tr$line
  )
}

p |>
  layout(
    updatemenus = list(list(
      type    = "dropdown",
      active  = 0,
      buttons = buttons,
      x       = 0,
      xanchor = "left",
      y       = 1.15,
      yanchor = "top"
    )),
    xaxis      = list(title = "Date"),
    yaxis      = list(title = "Weekly Cases", tickformat = ","),
    showlegend = FALSE,
    margin     = list(t = 80)
  )
```

# Data

```{r}
#| title: "Sri Lanka Weekly Dengue Surveillance Records"
#| message: false
#| warning: false

sl |>
  select(epi_week, district, cases) |>
  rename(
    `Epidemiological Week` = epi_week,
    District               = district,
    `Reported Cases`       = cases
  ) |>
  datatable(
    filter   = "top",
    rownames = FALSE,
    options  = list(
      pageLength = 15,
      scrollX    = TRUE,
      dom        = "Bfrtip",
      buttons    = c("copy", "csv", "excel")
    ),
    extensions = "Buttons"
  )
```