---
title: "Exploring Dengue in Sri Lanka"
author: "AS2022434"
format: dashboard
editor: source
theme: flatly
---

```{r}
#| label: load-packages
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(plotly)
library(dygraphs)
library(xts)
library(tsibble)
library(feasts)
library(fabletools)
library(timetk)
library(slider)
library(DT)
library(leaflet)
library(crosstalk)
library(denguedatahub)
```

```{r}
#| label: load-data
#| message: false
#| warning: false

data(srilanka_weekly_data)
```

```{r}
#| label: clean-data
#| message: false
#| warning: false

# Clean Dates and Standardize
sl <- srilanka_weekly_data |>
  mutate(start.date = parse_date_time(start.date, orders = c("mdy", "ymd"))) |>
  mutate(start.date = as.Date(start.date)) |>
  arrange(district, start.date)

# National Totals
sl_total <- sl |>
  group_by(start.date) |>
  summarise(cases = sum(cases, na.rm = TRUE), .groups = "drop")
sl_xts <- xts(sl_total$cases, order.by = sl_total$start.date)

# Annual & Top Districts
annual_district <- sl |>
  mutate(year = year(start.date)) |>
  group_by(district, year) |>
  summarise(annual_cases = sum(cases, na.rm = TRUE), .groups = "drop")

top10 <- annual_district |>
  group_by(district) |>
  summarise(total = sum(annual_cases), .groups = "drop") |>
  slice_max(total, n = 10) |>
  pull(district)

# Map Coordinates
district_coords <- tibble(
  district = c("Colombo", "Gampaha", "Kalutara", "Kandy", "Matale",
               "Nuwara Eliya", "Galle", "Matara", "Hambantota",
               "Jaffna", "Kilinochchi", "Mannar", "Vavuniya", "Mullaitivu",
               "Batticaloa", "Ampara", "Kalmunai", "Trincomalee",
               "Kurunegala", "Puttalam", "Anuradhapura", "Polonnaruwa",
               "Badulla", "Monaragala", "Ratnapura", "Kegalle"),
  lat = c(6.927, 7.084, 6.585, 7.291, 7.470, 6.970, 6.054, 5.948, 6.124,
          9.661, 9.395, 8.978, 8.751, 9.267, 7.717, 7.284, 7.414, 8.587,
          7.486, 8.036, 8.335, 7.939, 6.989, 6.873, 6.680, 7.253),
  lon = c(79.861, 80.007, 79.960, 80.636, 80.624, 80.783, 80.221, 80.536, 81.119,
          80.025, 80.399, 79.904, 80.497, 80.812, 81.700, 81.672, 81.836, 81.234,
          80.362, 79.840, 80.406, 81.000, 81.056, 81.349, 80.399, 80.349)
)
yearly_map_data <- left_join(district_coords, annual_district, by = "district")

# Crosstalk SharedData
shared_sl <- SharedData$new(sl)

# Pre-calculate STL for Top 4 to put in tabs
calc_stl <- function(dist_name) {
  sl |> 
    filter(district == dist_name) |>
    mutate(year_week = yearweek(start.date)) |>
    arrange(year_week) |>
    as_tsibble(index = year_week) |>
    model(STL(cases ~ season(period = 52))) |>
    components() |>
    autoplot() + theme_bw(base_size = 10) + labs(x = NULL, title = paste("STL -", dist_name))
}

stl_colombo <- calc_stl("Colombo")
stl_gampaha <- calc_stl("Gampaha")
stl_kalutara <- calc_stl("Kalutara")
stl_kandy <- calc_stl("Kandy")

# Pre-calculate Heatmaps for Top 4
calc_heat <- function(dist_name) {
  sl |> filter(district == dist_name) |>
    mutate(Year = fct_rev(factor(year(start.date))), Week = isoweek(start.date)) |>
    ggplot(aes(x = Week, y = Year, fill = cases)) +
    geom_tile(colour = "white", linewidth = 0.2) +
    scale_fill_viridis_c(option = "inferno", name = "Cases") +
    labs(x = "Week of Year", y = "Year", title = paste("Heatmap -", dist_name)) +
    theme_minimal(base_size = 10)
}

heat_colombo <- calc_heat("Colombo")
heat_gampaha <- calc_heat("Gampaha")
heat_kalutara <- calc_heat("Kalutara")
heat_kandy <- calc_heat("Kandy")
```

# Overview

## Row {height="20%"}

```{r}
#| content: valuebox
#| title: "Total Cases (All Time)"
list(icon = "virus", color = "danger", value = format(sum(sl_total$cases, na.rm = TRUE), big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Peak Weekly Cases"
peak <- sl_total |> slice_max(cases, n = 1, with_ties = FALSE)
list(icon = "arrow-up-circle", color = "warning", value = format(peak$cases, big.mark = ","))
```

```{r}
#| content: valuebox
#| title: "Peak Week"
list(icon = "calendar-week", color = "info", value = paste0("Week ", isoweek(peak$start.date), ", ", year(peak$start.date)))
```

```{r}
#| content: valuebox
#| title: "Districts Covered"
list(icon = "map", color = "success", value = n_distinct(sl$district))
```

## Row {height="80%"}

### Column {width="60%"}

```{r}
#| title: "Weekly Dengue Cases — All Sri Lanka"
dygraph(sl_xts) |>
  dyRangeSelector() |>
  dyOptions(strokeWidth = 1.5, fillGraph = TRUE, fillAlpha = 0.12, colors = "#c0392b") |>
  dyHighlight(highlightSeriesOpts = list(strokeWidth = 2.5))
```

### Column {width="40%"}

```{r}
#| title: "Annual Cases by District (Top 10)"
p_top10 <- annual_district |>
  filter(district %in% top10) |>
  ggplot(aes(x = year, y = annual_cases, colour = district, group = district)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  scale_colour_brewer(palette = "Paired") +
  labs(x = "Year", y = "Annual Cases", colour = "District") +
  theme_bw(base_size = 11)

ggplotly(p_top10) |> layout(hovermode = "x unified")
```

# District Map

## Row {.tabset}

### Map by Year

```{r}
#| title: "Dengue Cases by District — Select Year"
#| message: false
#| warning: false

max_cases <- max(yearly_map_data$annual_cases, na.rm = TRUE)
pal <- colorNumeric(palette = "Reds", domain = c(0, max_cases))
years_list <- sort(unique(yearly_map_data$year))

m <- leaflet() |>
  addTiles() |>
  setView(lng = 80.7, lat = 7.8, zoom = 7)

for (yr in years_list) {
  d <- yearly_map_data |> filter(year == yr)
  m <- m |>
    addCircleMarkers(
      data        = d,
      lng         = ~lon,
      lat         = ~lat,
      radius      = ~pmax(sqrt(annual_cases) / 10, 3),
      color       = ~pal(annual_cases),
      fillColor   = ~pal(annual_cases),
      fillOpacity = 0.85,
      opacity     = 1,
      popup       = ~paste0("<b>District:</b> ", district, "<br><b>Year:</b> ", year, "<br><b>Cases:</b> ", format(annual_cases, big.mark = ",")),
      label       = ~district,
      group       = as.character(yr)
    )
}

m |>
  addLayersControl(
    baseGroups = as.character(years_list),
    options    = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(pal = pal, values = c(0, max_cases), title = "Annual Cases")
```

### Cases by District

```{r}
#| title: "Annual Cases by District — Animated by Year"
yearly_map_data |>
  arrange(annual_cases) |>
  plot_ly(
    x           = ~annual_cases,
    y           = ~district,
    frame       = ~year,
    type        = "bar",
    orientation = "h",
    marker      = list(color = ~annual_cases, colorscale = "Reds"),
    text        = ~paste0("District: ", district, "<br>Year: ", year, "<br>Cases: ", format(annual_cases, big.mark = ",")),
    hoverinfo   = "text"
  ) |>
  layout(
    xaxis      = list(title = "Annual Cases", tickfont = list(size = 13)),
    yaxis      = list(title = "", tickfont = list(size = 13)),
    margin     = list(l = 120, r = 20, t = 30, b = 50),
    showlegend = FALSE,
    height     = 600
  ) |>
  animation_opts(frame = 800, transition = 400) |>
  animation_slider(currentvalue = list(prefix = "Year: "))
```

# Interactive Trends

## Sidebar {.sidebar}

```{r}
# Client-side JavaScript dropdown! No Shiny required.
filter_select(
  id = "district_chooser",
  label = "Choose a District to isolate its Time Series:",
  sharedData = shared_sl,
  group = ~district,
  multiple = FALSE
)
```

---

*Select a district from the dropdown above. The Time Series chart to the right and the Data table on the 'Data' page will instantly update.*

## Row

```{r}
#| title: "Weekly Cases — Filtered by Sidebar Dropdown"

plot_ly(shared_sl, x = ~start.date, y = ~cases, color = ~district, 
        colors = "viridis", type = 'scatter', mode = 'lines',
        text = ~paste("Date:", start.date, "<br>Cases:", cases),
        hoverinfo = "text") |>
  layout(
    xaxis = list(title = "Date"),
    yaxis = list(title = "Cases"),
    showlegend = FALSE
  )
```

# Deep Dive: Top 4 Districts

## Row {.tabset}

### Colombo

```{r}
#| title: "Colombo: STL & Heatmap"
#| fig-height: 8
subplot(
  ggplotly(stl_colombo), 
  ggplotly(heat_colombo, tooltip = c("x", "y", "fill")) |> layout(hovermode = "closest"), 
  nrows = 2, margin = 0.05, titleY = TRUE
)
```

### Gampaha

```{r}
#| title: "Gampaha: STL & Heatmap"
#| fig-height: 8
subplot(
  ggplotly(stl_gampaha), 
  ggplotly(heat_gampaha, tooltip = c("x", "y", "fill")) |> layout(hovermode = "closest"), 
  nrows = 2, margin = 0.05, titleY = TRUE
)
```

### Kalutara

```{r}
#| title: "Kalutara: STL & Heatmap"
#| fig-height: 8
subplot(
  ggplotly(stl_kalutara), 
  ggplotly(heat_kalutara, tooltip = c("x", "y", "fill")) |> layout(hovermode = "closest"), 
  nrows = 2, margin = 0.05, titleY = TRUE
)
```

### Kandy

```{r}
#| title: "Kandy: STL & Heatmap"
#| fig-height: 8
subplot(
  ggplotly(stl_kandy), 
  ggplotly(heat_kandy, tooltip = c("x", "y", "fill")) |> layout(hovermode = "closest"), 
  nrows = 2, margin = 0.05, titleY = TRUE
)
```

# National Anomalies

## Row {height="40%"}

```{r}
#| title: "52-Week Rolling Mean — All Sri Lanka"
p_roll <- sl_total |>
  mutate(MA52 = slide_dbl(cases, mean, .before = 51, .complete = TRUE)) |>
  ggplot(aes(x = start.date)) +
  geom_line(aes(y = cases), colour = "grey70", linewidth = 0.4, alpha = 0.8) +
  geom_line(aes(y = MA52), colour = "#c0392b", linewidth = 1) +
  labs(x = "Date", y = "Cases", caption = "Red = 52-week rolling mean | Grey = weekly observed") +
  theme_bw(base_size = 11)

ggplotly(p_roll) |> layout(hovermode = "x unified")
```

## Row {height="60%"}

```{r}
#| title: "Anomaly Detection — Highest Burden Districts"
#| message: false
#| warning: false

top4_data <- sl |> filter(district %in% c("Colombo", "Gampaha", "Kalutara", "Kandy"))

top4_data |>
  group_by(district) |>
  plot_anomaly_diagnostics(
    .date_var    = start.date, 
    .value       = cases, 
    .facet_ncol  = 2, 
    .interactive = TRUE,
    .title       = ""
  )
```

# Data

```{r}
#| title: "Searchable Dataset (Responds to Sidebar Filter on 'Interactive Trends' page)"
datatable(
  shared_sl,
  filter  = "top",
  options = list(pageLength = 15, scrollX = TRUE),
  rownames = FALSE,
  colnames = c("Date", "District", "Cases")
)
```