---
title: "Exploring Dengue in Sri Lanka"
author: "Shenan Nathaniel"
format: dashboard
editor: source
theme: flatly
---

```{r}
#| label: load-packages
#| message: false
#| warning: false

library(tidyverse)
library(lubridate)
library(plotly)
library(dygraphs)
library(xts)
library(tsibble)
library(feasts)
library(fabletools)
library(slider)
library(DT)
library(leaflet)
library(crosstalk)
library(htmltools)
library(denguedatahub)
```

```{r}
#| label: load-data
#| message: false
#| warning: false

data(srilanka_weekly_data)
```

```{r}
#| label: clean-data
#| message: false
#| warning: false

# Clean dates, standardize timelines, and create a logical Epidemiological Week label
sl <- srilanka_weekly_data |>
  mutate(start.date = parse_date_time(start.date, orders = c("mdy", "ymd"))) |>
  mutate(start.date = as.Date(start.date)) |>
  mutate(epi_week = paste0(year(start.date), " - W", sprintf("%02d", isoweek(start.date)))) |>
  arrange(district, start.date)

# National Totals
sl_total <- sl |>
  group_by(start.date) |>
  summarise(
    cases = sum(cases, na.rm = TRUE), 
    epi_week = first(epi_week),
    .groups = "drop"
  )
sl_xts <- xts(sl_total$cases, order.by = sl_total$start.date)

# Annual Totals by District
annual_district <- sl |>
  mutate(year = year(start.date)) |>
  group_by(district, year) |>
  summarise(annual_cases = sum(cases, na.rm = TRUE), .groups = "drop")

# Identify top burden regions for the line graphs
top10 <- annual_district |>
  group_by(district) |>
  summarise(total = sum(annual_cases), .groups = "drop") |>
  slice_max(total, n = 10) |>
  pull(district)

# Geographic Coordinates
district_coords <- tibble(
  district = c("Colombo", "Gampaha", "Kalutara", "Kandy", "Matale",
               "Nuwara Eliya", "Galle", "Matara", "Hambantota",
               "Jaffna", "Kilinochchi", "Mannar", "Vavuniya", "Mullaitivu",
               "Batticaloa", "Ampara", "Kalmunai", "Trincomalee",
               "Kurunegala", "Puttalam", "Anuradhapura", "Polonnaruwa",
               "Badulla", "Monaragala", "Ratnapura", "Kegalle"),
  lat = c(6.927, 7.084, 6.585, 7.291, 7.470, 6.970, 6.054, 5.948, 6.124,
          9.661, 9.395, 8.978, 8.751, 9.267, 7.717, 7.284, 7.414, 8.587,
          7.486, 8.036, 8.335, 7.939, 6.989, 6.873, 6.680, 7.253),
  lon = c(79.861, 80.007, 79.960, 80.636, 80.624, 80.783, 80.221, 80.536, 81.119,
          80.025, 80.399, 79.904, 80.497, 80.812, 81.700, 81.672, 81.836, 81.234,
          80.362, 79.840, 80.406, 81.000, 81.056, 81.349, 80.399, 80.349)
)
yearly_map_data <- left_join(district_coords, annual_district, by = "district")

# Isolate page-specific crosstalk datasets to stop cross-page communications
trends_shared <- SharedData$new(sl, group = "trends_page")
```

# Overview

## Row {height="35%"}

```{r}
#| content: valuebox
#| title: "Total Cases (All Time)"
list(
  icon  = "virus",
  color = "danger",
  value = format(sum(sl_total$cases, na.rm = TRUE), big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Peak Weekly Cases"
peak <- sl_total |> slice_max(cases, n = 1, with_ties = FALSE)
list(
  icon  = "arrow-up-circle",
  color = "warning",
  value = format(peak$cases, big.mark = ",")
)
```

```{r}
#| content: valuebox
#| title: "Peak Week"
list(
  icon  = "calendar-week",
  color = "info",
  value = peak$epi_week
)
```

## Row {height="65%"}

### Column {width="100%"}

### Core Exploratory Findings & Metadata Summary

Welcome to the Dengue Surveillance Dashboard for Sri Lanka. This analytical platform processes structural epidemiological records across all administrative districts to track outbreaks, evaluate seasonal distribution changes, and isolate historical trends.

#### Key Insights Captured:
* **The High-Burden Corridor:** Preliminary analysis confirms that the Western Province (consisting of Colombo, Gampaha, and Kalutara) bears the highest historical caseload in the country. 
* **Temporal Outbreak Dynamics:** The country experienced its highest all-time epidemiological transmission threshold during the year 2017, driven by heavy monsoon-associated vector proliferation.
* **Logical Structuring:** All temporal metrics within this platform have been converted to formal Epidemiological Weeks (Epi Weeks) rather than single days, matching standard public health reporting conventions.

*Navigate through the top tab structure to inspect national trend lines, geographic distributions, individual district breakdowns, and clear database records.*

# National Trends

## Row

### Column {width="60%"}

```{r}
#| title: "Weekly Transmission Volume — National Total Timeline"
dygraph(sl_xts) |>
  dyRangeSelector() |>
  dyOptions(
    strokeWidth = 1.5,
    fillGraph   = TRUE,
    fillAlpha   = 0.12,
    colors      = "#c0392b"
  ) |>
  dyHighlight(highlightSeriesOpts = list(strokeWidth = 2.5))
```

### Column {width="40%"}

```{r}
#| title: "Annual Outbreak Aggregations (Top 10 Highest-Burden Districts)"
p_top10 <- annual_district |>
  filter(district %in% top10) |>
  ggplot(aes(x = year, y = annual_cases, colour = district, group = district,
             text = paste0("District: ", district, "<br>Year: ", year, "<br>Annual Cases: ", format(annual_cases, big.mark = ",")))) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.5) +
  scale_colour_brewer(palette = "Paired") +
  labs(x = "Year", y = "Annual Cases", colour = "District") +
  theme_bw(base_size = 11)

ggplotly(p_top10, tooltip = "text") |> 
  layout(hovermode = "closest")
```

# District Map

## Row {.tabset}

### Map by Year

```{r}
#| title: "Geographic Spread by Selected Year"
#| message: false
#| warning: false

max_cases <- max(yearly_map_data$annual_cases, na.rm = TRUE)
pal <- colorNumeric(palette = "Reds", domain = c(0, max_cases))
years_list <- sort(unique(yearly_map_data$year))

m <- leaflet() |>
  addTiles() |>
  setView(lng = 80.7, lat = 7.8, zoom = 7)

for (yr in years_list) {
  d <- yearly_map_data |> filter(year == yr)
  m <- m |>
    addCircleMarkers(
      data        = d,
      lng         = ~lon,
      lat         = ~lat,
      radius      = ~pmax(sqrt(annual_cases) / 10, 3),
      color       = ~pal(annual_cases),
      fillColor   = ~pal(annual_cases),
      fillOpacity = 0.85,
      opacity     = 1,
      popup       = ~paste0("<b>District:</b> ", district,
                            "<br><b>Year:</b> ", year,
                            "<br><b>Annual Cases:</b> ", format(annual_cases, big.mark = ",")),
      label       = ~district,
      group       = as.character(yr)
    )
}

# Position explicitly moved to "bottomleft" to fix visibility issues
m |>
  addLayersControl(
    baseGroups = as.character(years_list),
    options    = layersControlOptions(collapsed = FALSE)
  ) |>
  addLegend(
    pal      = pal,
    values   = c(0, max_cases),
    title    = "Annual Cases",
    position = "bottomleft"
  )
```

### Annual Case Matrix (All 26 Districts)

```{r}
#| title: "Comparative District Timelines (Replaces Broken Animation)"
p_matrix <- ggplot(annual_district, aes(x = year, y = reorder(district, annual_cases, sum), fill = annual_cases,
                        text = paste0("District: ", district, "<br>Year: ", year, "<br>Cases: ", format(annual_cases, big.mark = ",")))) +
  geom_tile(color = "white", linewidth = 0.1) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1, name = "Cases", labels = scales::comma) +
  labs(x = "Reporting Year", y = "") +
  theme_minimal(base_size = 11)

ggplotly(p_matrix, tooltip = "text") |> 
  layout(margin = list(l = 100, r = 20, t = 30, b = 50))
```

# Interactive Trends

## Sidebar {.sidebar}

```{r}
filter_select(
  id         = "trends_district_chooser",
  label      = "Isolate Specific District Line:",
  sharedData = trends_shared,
  group      = ~district,
  multiple   = FALSE
)
```

---

*Use the isolated drop-down menu above to select a single district line. This page functions independently to eliminate cross-widget clutter.*

## Row

```{r}
#| title: "Weekly Trajectory Analysis by Selected District"

plot_ly(trends_shared, x = ~start.date, y = ~cases, color = ~district, 
        colors = "Set2", type = 'scatter', mode = 'lines',
        text = ~paste0("District: ", district, "<br>Epi Week: ", epi_week, "<br>Weekly Cases: ", cases),
        hoverinfo = "text") |>
  layout(
    xaxis = list(title = "Timeline (Zoomable Grid)"),
    yaxis = list(title = "Observed Cases"),
    showlegend = TRUE
  )
```

# District Deep Dive

## Row {.tabset}

```{r}
#| results: asis
#| echo: false

all_districts <- sort(unique(sl$district))

for (i in seq_along(all_districts)) {
  current_dist <- all_districts[i]
  
  cat(paste0("### ", current_dist, "\n\n"))
  
  # Filter data for specific iteration
  dist_data <- sl |> filter(district == current_dist)
  
  # 1. STL Decomposition Calculation
  dist_ts <- dist_data |>
    mutate(year_week = yearweek(start.date)) |>
    distinct(year_week, .keep_all = TRUE) |>
    arrange(year_week) |>
    as_tsibble(index = year_week)
  
  stl_model <- dist_ts |>
    model(STL(cases ~ season(period = 52))) |>
    components()
  
  p_stl <- autoplot(stl_model) + 
    theme_bw(base_size = 9) + 
    labs(x = NULL, title = paste(current_dist, "— Temporal Dissection (Trend & Seasonality)"))
  
  # 2. High-Contrast Yellow-Orange-Red Heatmap
  p_heat <- dist_data |>
    mutate(Year = fct_rev(factor(year(start.date))), Week = isoweek(start.date)) |>
    ggplot(aes(x = Week, y = Year, fill = cases,
               text = paste0("Year: ", year(start.date), "<br>Week: ", Week, "<br>Cases: ", cases))) +
    geom_tile(colour = "white", linewidth = 0.1) +
    scale_fill_viridis_c(option = "YlOrRd", name = "Cases") +
    labs(x = "Calendar Week of Year", y = "Year", title = paste(current_dist, "— Outbreak Grid Density")) +
    theme_minimal(base_size = 9)
  
  # Stitch components side-by-side using an elegant dynamic subplot widget structure
  sub_plot <- subplot(
    ggplotly(p_stl), 
    ggplotly(p_heat, tooltip = "text") |> layout(hovermode = "closest"), 
    nrows = 2, margin = 0.07, titleY = TRUE
  )
  
  print(htmltools::tagList(sub_plot))
  cat("\n\n")
}
```

# Data

```{r}
#| title: "Independent Data Repository Browser"

# Standard dataframe passed directly to isolate search metrics and stop background cross-filtering
sl |>
  select(epi_week, district, cases) |>
  rename(`Epidemiological Week` = epi_week, District = district, `Registered Cases` = cases) |>
  datatable(
    filter   = "top",
    rownames = FALSE,
    options  = list(pageLength = 15, scrollX = TRUE)
  )
```